---
title: "Species Extinctions"
author: "Grace Pratt, Sarah Tang"
maketitle: true
output: github_document
---

```{r include=FALSE}
library("tidyverse")
library("httr")
library("jsonlite")
library(stringr)
knitr::opts_chunk$set(comment=NA)
```


## Extinctions Module

_Are we experiencing the sixth great extinction?_  

What is the current pace of extinction? Is it accelerating? How does it compare to background extinction rates?

## Computational Topics

In order to assess the current rate of extinction, we:

- Access data from a RESTful API
- Handle errors 
- Use the JSON data format
- Implement regular expressions
- Work with missing values


## CURL and REST

We use the `httr` package to make a single API query for the species Acaena exigua.

```{r}
genus <- "Acaena"
species <- "exigua"
api_call <- paste0("http://api.iucnredlist.org/index/species/", genus, "-", species, ".json")
resp <- GET(api_call)
resp
```

*By examining the response and the content of the response, we can determine that the call was successful because the status is 200. The return type of the object is a json object. Later, we parse the returned object into test and then present the data as a data frame.*

# Calculating Extinction Rates: Putting it all together

First, we download a data file containing a list of extinct species names to know how to make queries to the IUCN REST API.


```{r message=FALSE}
extinct = read_csv("https://espm-157.github.io/extinction-module/extinct.csv")
```


Now, we write a function to extract the rationale for the extinction for all extinct species in the data set (see above file).

```{r}
api_fn <- function(genus, species) {
  api_call <- paste0("http://api.iucnredlist.org/index/species/", genus, "-", species, ".json")
  Sys.sleep(0.1)
  resp <- GET(api_call)
  status <- httr::status_code(resp)

  if (status > 300) {
    return(data_frame(Genus = genus, Species = species, rationale = NA, status_code = status))
  } 
  
  out <- content(resp, as = "text")
  df <- jsonlite::fromJSON(out)
  rationale <- df$rationale
  
  data_frame(Genus = genus, Species = species, API = api_call, rationale = rationale, status_code = status)
}

df <- map2_dfr(extinct$Genus, extinct$Species, api_fn)
```

Now we create a function that can extract the date from the rationale. To do this we make some assumptions. We first assume that any 4 digit number in the rationale is the year the species went extinct. We also assume that any number between 21 - 1000 is the "time since" the species went extinct and that any 2 digit number bellow 21 is the century that the the species went extinct. For centuries, all dates are the first year of that century (e.g. 20th century became 1900).

```{r}
time_since <- function(x) { #assuming any number between 21 - 1000 is the "time since" the species went extinct
  if_else((x > 21 & x < 1000), (2018 - x), x)
}

century <- function(x) { #assuming any 2 digit number bellow 21 is the century that the the species went extinct
  if_else((x <= 21), ((x - 1) * 100), x)
}

txt <- df$rationale
dates <- as.numeric(stringr::str_extract(txt, "(\\d{2,})"))
df_years <- df %>% mutate(extracted_dates = dates) %>%
  mutate(Year = time_since(century(extracted_dates))) 
df_years
```


## Histogram of Extinction Dates

We can get a sense for the tempo of extinctions by plotting extinctions since 1500 in 25-year interval bins.  

```{r}
df_extinct <- df_years %>% filter(!is.na(Year)) %>%
  filter(Year > 1500)

ggplot(df_extinct, aes(x = Year)) + 
  geom_histogram(breaks = seq(1500, 2025, by = 25)) + 
  labs(title = "Extinctions since 1500 per 25 years", y = "Frequency")
```

# Extinctions by group

In order to see how extinctions vary across different species, we compute the number of extinctions from 1500 - 1900 and from 1900 to present of each of the following taxonomic groups: 

  - Vertebrates
  - Mammals
  - Birds
  - Fish
  - Amphibians
  - Reptiles
  - Insects
  - Plants
 
First we create dataframes for each taxanomic group.
```{r message=FALSE}
mammals <- extinct %>%
  filter(Class == "MAMMALIA") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()

birds <- extinct %>%
  filter(Class == "AVES") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()

fish <- extinct %>%
  filter(Class == "ACTINOPTERYGII" 
         | Class == "CEPHALASPIDOMORPHI"
         | Class == "MYXINI"
         | Class == "ELASMOBRANCHII"
         | Class == "HOLOCEPHALI"
         | Class == "SACROPTERYGII") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()

amphibians <- extinct %>%
  filter(Class == "AMPHIBIA") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()
  
reptiles <- extinct %>%
  filter(Class == "REPTILIA") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()

insects <- extinct %>%
  filter(Class == "INSECTA") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()

plants <- extinct %>%
  filter(Kingdom == "PLANTAE") %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()

vertebrates <- mammals %>%
  rbind(birds, amphibians, reptiles, fish) %>%
  left_join(df_years) %>%
  select(Genus, Species, rationale, Year) %>%
  na.omit()
```


Now we filter to get the number of species for each group that went extict between 1500-1900 and post-1900. 
```{r}
between_1500_1900 <- function(df){
  df %>%
    filter(Year >= 1500, Year < 1900) %>%
    nrow()
}

post_1900 <- function(df){
  df %>% 
    filter(Year >= 1900) %>%
    nrow()
}

taxonomy_vector <- c("mammals", "birds", "fish", "amphibians", "reptiles", "insects", "plants", "vertebrates")

our_extinctions <- data_frame("Taxonomic Group" = taxonomy_vector, 
           "Extinctions 1500-1900" = 
             c(between_1500_1900(mammals),between_1500_1900(birds), between_1500_1900(fish), between_1500_1900(amphibians),between_1500_1900(reptiles), between_1500_1900(vertebrates), between_1500_1900(plants), between_1500_1900(vertebrates)),
           "Extinctions post-1900" = 
             c(post_1900(mammals), post_1900(birds), post_1900(fish),post_1900(amphibians), post_1900(reptiles), post_1900(insects), post_1900(plants), post_1900(vertebrates)))

our_extinctions
```


In order to see how accurate our estimates are, we compare our estimates to Table 1 of [Ceballos et al (2015)](http://doi.org/10.1126/sciadv.1400253).  

```{r}
table1 <- data_frame("Taxonomic Group" = c("mammals", "birds", "fish", "amphibians", "reptiles", "vertebrates"), 
           "Extinctions 1500-1900" = c(42, 83, 0, 2, 13, 140),
           "Extinctions post-1900" = c(35, 57, 66, 32, 8, 198))

comparison <- our_extinctions %>%
  filter(`Taxonomic Group` %in% table1$`Taxonomic Group`) %>%
  mutate("Difference 1500-1900" = `Extinctions 1500-1900` - table1$`Extinctions 1500-1900`)%>%
  mutate("Difference post-1900" = `Extinctions post-1900` - table1$`Extinctions post-1900`)
  
comparison
```


*Overall, the numbers of extinct species we estimated was less than the numbers of extinctions estimated by Ceballos et al. The only exception to this was the number of fish that went extinct between 1500 and 1900, where we estimated 4 more extinct species than Ceballos et al. This could be due to our taxanomic defintiion of fish, which may be different from the one used by Ceballos et al.*

## Weighing by number of species

The number of species going extinct per century in a given taxonomic group will be influenced by how many species are present in the group to begin with. Overall, these numbers do not change greatly over a period of a few hundred years, so we were able to make the relative comparisons between the roughly pre-industrial and post-industrial periods above.  

As discussed by Tony Barnosky in the introductory video (or in [Ceballos et al (2015)](http://doi.org/10.1126/sciadv.1400253) paper), if we want to compare these extinction rates against the long-term palentological record, it is necessary to weigh the rates by the total number of species. That is, to compute the number of extinctions per million species per year (MSY; equivalently, the number extinctions per 10,000 species per 100 years).  

A) First, we will compute how many species are present in each of the taxonomic groups.  To do so, we need a table that has not only extinct species, but all assessed species.  We will once again query this information from the IUCN API.

```{r}
api_calls <- map((0:10), function(page_num){paste0("http://apiv3.iucnredlist.org/api/v3/species/page/", page_num, "?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee")})

resp2 <- map(api_calls[1:10], function(url){
  Sys.sleep(0.5)
  GET(url)
})

out <- map(resp2, content, as = "text")
df_page <- map(out, fromJSON)

all_df <- plyr::ldply(df_page, data.frame)

```


B) Based on the complete data, we count the number of species in each group. Then we use these numbers to compute MSY, the number extinctions per 10,000 species per 100 years, for each of the taxanomic groups.

```{r}
allmammals <- all_df %>%
  filter(result.class_name == "MAMMALIA") %>%
  nrow()

allbirds <- all_df %>%
  filter(result.class_name == "AVES") %>%
  nrow()

allfish <- all_df %>%
  filter(result.class_name == "ACTINOPTERYGII" 
         | result.class_name == "CEPHALASPIDOMORPHI"
         | result.class_name == "MYXINI"
         | result.class_name == "ELASMOBRANCHII"
         | result.class_name == "HOLOCEPHALI"
         | result.class_name == "SACROPTERYGII") %>%
  nrow()

allamphibians <- all_df %>%
  filter(result.class_name == "AMPHIBIA") %>%
  nrow()

allreptiles <- all_df %>%
  filter(result.class_name == "REPTILIA") %>%
  nrow()

allinsects <- all_df %>%
  filter(result.class_name == "INSECTA") %>%
  nrow()

allplants <- all_df %>%
  filter(result.kingdom_name == "PLANTAE") %>%
  nrow()

allvertebrates <- allmammals %>%
  sum(allbirds, allamphibians, allreptiles, allfish)

msy_vector <- c(nrow(mammals)/(allmammals/10000)/100,
  nrow(birds)/(allbirds/10000)/100,
  nrow(fish)/(allfish/10000)/100,
  nrow(amphibians)/(allamphibians/10000)/100,
  nrow(reptiles)/(allreptiles/10000)/100,
  nrow(insects)/(allinsects/10000)/100,
  nrow(plants)/(allplants/10000)/100,
  nrow(vertebrates)/(allvertebrates/10000)/100)

msy_df <- data_frame("taxanomic_group" = taxonomy_vector, "MSY" = msy_vector)
msy_df

```

*Overall, we can see that our estimates for every group are less than the overall historical average of about 2 MSY. This is likely due to our limited algorithm that could not extract dates for all of the extinct species. Therefore, there are likely many species that went extinct after 1500 that we aren't including in our analysis. This makes our MSY artificially low. Looking at the MSYs calculated by Ceballos et al, we can see that our MSY estimates should be increased by 4-13 points depending on the taxanomic group.*

## Improving our algorithm

*In parsing the data with regular expressions and focusing only on four digit values, we encountered certain data that resulted in missing values. Our data extraction failed because we were only extracting years and not accounting for instances when there were 2 or 3 digits. Some main cases that we identified were in rationale when the date depicted was a century (ex: 16th century) or when the corresponding number referred to the number of years ago that a species went extinct (ex: 70 years ago).*

*Ultimately, we went back through our algorithm and modified how we handled regular expressions to reduce the number of missing values. We considered the two cases mentioned above and changed our algorithm to include regular expressions to grab all 2, 3, and 4 digit numbers. With two digit numbers less than or equal to 21, we subtracted 1 and then mulitiplied by 100 to get the correct century, a ballpark estimate. For example, "16th century" would give us 1500. For numbers greater than 21 and less than 1000, as they clearly can't describe a century, we assumed them to be x years ago. We outputted 2018 - x for such numbers.*